---
title: "fmm.Rmd"
author: "Virgilio Gonzenbach"
date: "1/28/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, message=FALSE}
library(dplyr)
library(mclust)
library(FactoMineR)
library(factoextra)
library(kableExtra)

chop_df = read.csv('data/Chop_merged_wideV_tp1.csv')
penn_df = read.csv("data/Penn_merged_wideV.csv")[,-1] %>% filter(group != "?") # 3 observations in '?' group

# Exclude columns with only zeros
get_zero_cols = function(df){
  num_zeros = df %>% sapply(function(x) x == 0) %>% colSums()
  prop_zeros = num_zeros / nrow(df)
  zero_cols = names(which(num_zeros == nrow(df)))
  return(zero_cols)
}

get_zero_cols = function(df){
  num_zeros = df %>% sapply(function(x) x == 0) %>% colSums()
  prop_zeros = num_zeros / nrow(chop_df)
  zero_cols = names(which(prop_zeros == 1))
  return(zero_cols)
}

chop_df = chop_df %>% select(-all_of(get_zero_cols(chop_df)))
penn_df = penn_df %>% select(-all_of(get_zero_cols(penn_df)))

# Set seed for reproducibility
seed=42
set.seed(seed)
```

## Sample sizes

```{r}
t(table(c(chop_df$group, penn_df$group))) %>% kable(format="html") %>% kable_styling("striped", full_width = TRUE)
```


## Latent Profile Analysis

Latent Profile Analysis, or Gaussian Mixture Models, is a model-based clustering approach aimed at finding distinct "profile"/clusters based on continuous data. In LPA, the distribution of observed scores is assumed to come from a number of distinct (Gaussian/normal) distributions. LPA estimates parameters (i.e. mean, variance, covariances) for the specified number of distributions $g$.  

## 1. a. LPA on first PCs from 22q sample

```{r}
chop_gscores = chop_df[, -c(1:7)]
resPCA_22q = PCA(chop_gscores, scale.unit = FALSE, graph = FALSE, ncp=length(chop_gscores))
fviz_screeplot(resPCA_22q)
```

```{r}
pc_df = data.frame(resPCA_22q$ind$coord[,1:4])
mod = mclust::Mclust(pc_df)
summary(mod)
plot(mod, what = "BIC")
```

```{r}
plot(mod, what = "classification")
```

```{r}
plot(mod, what = "uncertainty")
```

```{r}
ICL = mclustICL(pc_df)
plot(ICL)
```

```{r}
LRT = mclustBootstrapLRT(pc_df, modelName = "VII")
LRT
```

Initialization?

### External Validation




## 2. a. PS group

Pre-processing: PCA applied directly on the PS gestalt scores (**not** projected onto 22q feature space).

Taking first 6 dimensions.

```{r}
demographics = c("bbl_id", "case_id", "sex", "race", "age", "age_at_photo")
ps_gscores = penn_df %>% filter(group != 'NC') %>% select(-c(all_of(demographics), group))

resPCA = PCA(ps_gscores, scale.unit = FALSE, graph = FALSE, ncp=length(chop_gscores))
fviz_screeplot(resPCA)
```

```{r}
pc_df = data.frame(resPCA$ind$coord[,1:6])
mod = Mclust(pc_df)
```

```{r}
summary(mod)
plot(mod, what = "BIC")
```
```{r}
plot(mod, what = "classification")
```

```{r}
plot(mod, what = "uncertainty")
```

```{r}
ICL = mclustICL(pc_df)
plot(ICL)
```
```{r}
LRT = mclustBootstrapLRT(pc_df, modelName = "VEI")
LRT
```