---
title: "fmm.Rmd"
author: "Virgilio Gonzenbach"
date: "1/28/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
```

```{r, message=FALSE}
library(tidyr)
library(dplyr)
library(mclust)
library(InPosition)
library(FactoMineR)
library(factoextra)
library(kableExtra)

chop_df = read.csv('data/Chop_merged_wideV_tp1.csv')
penn_df = read.csv("data/Penn_merged_wideV.csv")[,-1] %>% filter(group != "?") # 3 observations in '?' group

# Exclude columns with only zeros
get_zero_cols = function(df){
  num_zeros = df %>% sapply(function(x) x == 0) %>% colSums()
  prop_zeros = num_zeros / nrow(df)
  zero_cols = names(which(num_zeros == nrow(df)))
  return(zero_cols)
}

get_zero_cols = function(df){
  num_zeros = df %>% sapply(function(x) x == 0) %>% colSums()
  prop_zeros = num_zeros / nrow(chop_df)
  zero_cols = names(which(prop_zeros == 1))
  return(zero_cols)
}

chop_df = chop_df %>% select(-all_of(get_zero_cols(chop_df)))
penn_df = penn_df %>% select(-all_of(get_zero_cols(penn_df)))

# Set seed for reproducibility
seed=42
set.seed(seed)
```

## Sample sizes

```{r}
t(table(c(chop_df$group, penn_df$group))) %>% kable(format="html") %>% kable_styling("striped", full_width = TRUE)
```


## Latent Profile Analysis

Latent Profile Analysis, or Gaussian Mixture Models, is a model-based clustering approach aimed at finding distinct "profile"/clusters based on continuous data. In LPA, the distribution of observed scores is assumed to come from a number of distinct (Gaussian/normal) distributions. LPA estimates parameters (i.e. mean, variance, covariances) for the specified number of distributions $g$.  

## 1. a. LPA on first PCs from 22q sample

```{r}
chop_gscores = chop_df[, -c(1:7)]
resPCA_22q = PCA(chop_gscores, scale.unit = FALSE, graph = FALSE, ncp=length(chop_gscores))
# resPCA_22q = epPCA.inference.battery(chop_gscores, scale = FALSE, graphs=FALSE)
fviz_screeplot(resPCA_22q)
```

```{r}
get_loadings = function(resPCA, dim=1){
  # Extract metrics from PCA object
  r = resPCA$var$coord[, dim] / sqrt(resPCA$eig[dim])
  # is equal to 
  return(r)
}

plot_loadings = function(resPCA, dim=1){
  # Plot loadings
  r = get_loadings(resPCA, dim)
  filter = r^2 > mean(r^2, na.rm = TRUE)
  
  # Prep data
  df = data.frame(var=names(r), cor = r, row.names = NULL)[filter, ] %>% drop_na()
  df = df[order(df$cor, decreasing = TRUE),]
  df$var = factor(df$var, levels = rev(df$var)) #order before plotting
  
  # Plot
  title = sprintf("PC%s loadings", dim)
  p = df %>% ggplot(aes(x=var, y=cor)) + geom_col() + coord_flip() + 
    xlab("Gestalt Score") + ylab("Loading") + ggtitle(title) + 
    theme_bw() + theme(axis.text.y = element_text(size=10))
  return(p)
}

plot_loadings(resPCA_22q, 1)
```

```{r}
pc_df = data.frame(resPCA_22q$ind$coord[,1:4])
mod = mclust::Mclust(pc_df)
summary(mod)
```

## Delete?

```{r}
plot(mod, what = "BIC")
```


```{r}
plot(mod, what = "classification")
```

```{r}
LRT = mclustBootstrapLRT(pc_df, modelName = "VII")
LRT
```

### External Validation

Sex: 

```{r}
chisq.test(mod$classification, chop_df$sex)
```

Age:

```{r}
t.test(chop_df$age[mod$classification == 1], chop_df$age[mod$classification == 2])
```

Race:

```{r}
chisq.test(mod$classification, chop_df$race)
```


## 2. a. PS group

Pre-processing: PCA applied directly on the PS gestalt scores (**not** projected onto 22q feature space).

Taking first 6 dimensions.

```{r}
demographics = c("bbl_id", "case_id", "sex", "race", "age", "age_at_photo")
ps_df = penn_df %>% filter(group != 'NC')
ps_gscores =  ps_df %>% select(-c(all_of(demographics), group))

resPCA = PCA(ps_gscores, scale.unit = FALSE, graph = FALSE, ncp=length(chop_gscores))
fviz_screeplot(resPCA)
```

```{r}
pc_df = data.frame(resPCA$ind$coord[,1:6])
mod = Mclust(pc_df)
```

```{r}
summary(mod)
plot(mod, what = "BIC")
```
```{r}
plot(mod, what = "classification")
```
```{r}
plot_loadings(resPCA, 2)
```

```{r}
LRT = mclustBootstrapLRT(pc_df, modelName = mod$modelName)
LRT
```


### External Validation

Concordance with

```{r}
chisq.test(mod$classification, penn_df$group[penn_df$group != 'NC'])
```

```{r}
adjustedRandIndex(mod$classification, ps_df$group)
```

Sex:
```{r}
chisq.test(mod$classification,  ps_df$sex)
```

Race: 

```{r}
chisq.test(mod$classification,  ps_df$race)
```

Age: 

```{r}
t.test(pc_df$Dim.2[ps_df$group == 'SZ'], pc_df$Dim.2[ps_df$group == 'CR'])
```


## 3. a PS group projected scores

```{r}
weigh_scores2 = function(df, resPCA){
  # which gestalt scores are missing from penn_df? i.e. PS and NC groups
  nohits = rownames(resPCA$var$coord)[which(!(rownames(resPCA$var$coord) %in% colnames(df)))] 
  df[, nohits] = 0 # set missing to zero to define columns
   
  scores.mat = as.matrix(df[, rownames(resPCA$var$coord)]) 
  weighted_score = scores.mat %*% resPCA$svd$V
  
  return(weighted_score)
}

# Project scores
ps_gscores = penn_df %>% filter(group != 'NC') %>% select(-c(all_of(demographics), group))
pc_df = weigh_scores2(ps_gscores, resPCA_22q)[,1:4] %>% data.frame()
colnames(pc_df) = paste('Dim', 1:4, sep='.')
```

```{r}
mod = mclust::Mclust(pc_df)
summary(mod)
```

```{r}
plot(mod$BIC)
```

```{r}

```



