---
title: "Predicting PS status with 22q characteristics"
author: "Virgilio Gonzenbach, Joelle Jee"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document:
    toc: true
    toc_float: True
    theme: 'spacelab'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      warning = FALSE,
                      message = FALSE,
                      cache = FALSE,
                      cache.path = 'cache/PSpred/', 
                      fig.path = "figures/PSpred/")
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
```


```{r load, include=FALSE}
library(dplyr)
library(tidyr)
library(FactoMineR)
library(factoextra)
library(pander)
library(pROC)
library(kableExtra)

dat = read.csv('data/penn_chop_tp1_f2g_emotrics_passed_clean_030822.csv',
               check.names = FALSE) 
sz_cr = read.csv('data/penn_szcr_chop_tp1_wideV_f2g_030822.csv')
sz_cr = sz_cr[sz_cr$group %in% c('TD', 'SZ', 'CR'), c('bbl_id', 'group')]
chop_df = dat[dat$group == '22q', ]
penn_df = dat[dat$group != '22q', ]

# Exclude columns with only zeros
get_zero_cols = function(df){
  num_zeros = df %>% sapply(function(x) x == 0) %>% colSums()
  prop_zeros = num_zeros / nrow(df)
  zero_cols = names(which(num_zeros == nrow(df)))
  return(zero_cols)
}

chop_df = chop_df %>% select(-all_of(get_zero_cols(chop_df)))
penn_df = penn_df %>% select(-all_of(get_zero_cols(penn_df)))

# Set seed for reproducibility
seed=42
set.seed(seed)
```

In this analysis, psychosis spectrum (PS) status is predicted based on 22q11.2 deletion syndrome (22q) characteristics. First, PCA is performed on the 22q sample to estimate the relative contributions, i.e. the loadings, of each gestalt score to the axes of greatest variability (PCs) in this group. These loadings are then used to compute "22q-like" factor scores for the TD and PS groups:

- ROC analysis shows **PC2 alone correctly predicts PS status 66% of the time**. 

- Additionally, Bartlett’s test shows that **variance from the schizophrenia (SZ) group differs from that of clinical risk (CR) on PC2**.

- On the 22q sample, PC2 scores were associated with Psychosis Symptoms, Emotrics and MMSE scores, as well as sex and age.

- Prediction of **Schizophrenia Status** remained strong after correcting out-of-sample PC2 scores (i.e. from non-22q groups) by the above covariates. Prediction of **CR status** weakened after correction. 

## 0. Sample sizes

Total sample sizes in 03/08/22 data and non-missing data per variable:  

```{r}
# Get total in extended data
vars= c("height", "weight", "GAF_C", "GAF_H", "MMSE", "VIQ", "PIQ", "FSIQ", "DSM Dx", "PS")  
n_df = data.frame(case_id = dat$case_id, 
                  #group = dat$group,
                  !is.na.data.frame(dat[,vars]),
                  check.names = FALSE)

n_df = merge(read.csv('data/penn_szcr_chop_tp1_wideV_f2g_030822.csv')[, c("case_id", "group")], n_df, by='case_id')
n_size = t(table(n_df$group))
n_df = n_df %>% select(-case_id) %>% group_by(group) %>% summarize_all(sum)
n_tbl = rbind(n_size, t(n_df[,-1]))
n_tbl = cbind(n_tbl, rowSums(n_tbl))
colnames(n_tbl) = c(t(n_df)[1,], "total") 
n_tbl = as.table(n_tbl)
n_tbl %>% kable(format="html") %>% kable_styling("striped", full_width = TRUE)
```

Note: Emotrics available for all records.

## 1. Examining characteristics of 22q sample with PCA

PCA was performed without scaling on the 22q sample alone.

### Scree Plot

```{r}
syndromes = colnames(chop_df)[which(colnames(chop_df) == '22q11.2 deletion syndrome'):ncol(chop_df)]
demographics = colnames(chop_df)[1:(which(colnames(chop_df) == 'Brow_Height_Right') - 1)]
chop_df = chop_df[, c(demographics, syndromes)]
resPCA = PCA(chop_df[, syndromes], 
             scale.unit = FALSE, 
             graph = FALSE, 
             ncp=length(chop_df[,syndromes]))

fviz_screeplot(resPCA)
```

Using the "elbow rule", subsequent analysis will focus on the first 4 PCs. 

### Loadings

Loadings for the first 2 PCs are plotted below.

```{r}
get_loadings = function(resPCA, dim=1){
  # Extract metrics from PCA object
  r = resPCA$var$coord[,dim] / sqrt(resPCA$eig[dim,1])
  
  # is equal to 
  return(r)
}

plot_loadings = function(resPCA, dim=1){
  # Plot loadings
  r = get_loadings(resPCA, dim)
  filter = r^2 > mean(r^2, na.rm = TRUE) # only include loadings with more than average contributions
  
  # Prep data
  df = data.frame(var=names(r), cor = r, row.names = NULL)[filter, ] %>% drop_na()
  df = df[order(df$cor, decreasing = TRUE),]
  df$var = factor(df$var, levels = rev(df$var)) #order before plotting
  
  # Plot
  title = sprintf("PC%s contributions", dim)
  p = df %>% ggplot(aes(x=var, y=cor)) + geom_col() + coord_flip() + 
    xlab("Gestalt Score") + ylab("Loading") + ggtitle(title) + 
    theme_bw() + theme(axis.text.y = element_text(size=10))
  return(p)
}
```

#### PC1

```{r, fig.width=12, fig.height=8}
plot_loadings(resPCA, dim=1)
```

#### PC2

```{r, fig.width=12, fig.height=8}
plot_loadings(resPCA, dim=2)
```

### Covariate effects on PC2 scores

#### Demographics

There were significant effects of sex and age on PC2 scores. 

```{r}
cov_df = data.frame(chop_df[, c("case_id", "sex", "age", "race")], 
                    PC1 = resPCA$ind$coord[, 1], 
                    PC2 = resPCA$ind$coord[, 2])
anova(lm(PC2 ~ sex + age + race, data=cov_df))
```

```{r}
cov_df %>% 
  ggplot(aes(x=PC1, y=PC2, color = sex)) + 
  geom_point() + 
  geom_point(data = cov_df %>% 
               select(PC1, PC2, sex) %>% 
               group_by(sex) %>%  
               summarize_all(mean), size = 4, shape = 17) + 
  scale_color_manual(values = c("red", "blue")) + 
  ggtitle("22q-like PC scores for 22q sample coded by sex")
```

```{r, message=FALSE, warning=FALSE}
cov_df %>% 
  ggplot(aes(x=age, y=PC2)) + 
  geom_point() + 
  geom_smooth(method='lm')
```

The remaining results on this section show effects on PC2 with and without adjusting for sex and age (see 'adj PC2').

#### Diagnosis

##### Psychosis

22q patients with psychotic symptoms had significantly higher PC2 scores than 22q patients without psychosis.

```{r}
report.t = function(res.t, x, y){
  #' Report t-test in apa format
  #'' TODO: to fully automate: print groups from data.name, bonus: add formatting/substutions of var.name as parameter
  #' Make this work for a single test (check input is list)

    cat(sprintf("\n - %s (*M* = %s, *SD* = %s) and %s (*M* = %s, *SD* = %s), %s", 
            x,
            round(mean(res.t$data$x), 2), round(sd(res.t$data$x), 2),
            y,
            round(mean(res.t$data$y), 2), round(sd(res.t$data$y), 2),
            apa::apa(res.t, print = FALSE)), 
        sep="\n")
}
```


```{r}
patho_df = dat[, c("case_id", "DSM Dx", "PS")] 

# Make factors
patho_df$`DSM Dx` = as.factor(recode(patho_df$`DSM Dx`, `1` = "Yes", `0` = "No"))
patho_df$PS = as.factor(recode(patho_df$PS, `1` = "Yes", `0` = "No"))
patho_df = merge(patho_df, cov_df, by="case_id") %>% drop_na()
```

###### PC2

```{r, results='asis'}
x = patho_df$PC2[patho_df$PS == "Yes"]
y = patho_df$PC2[patho_df$PS == "No"]
report.t(apa::t_test(x, y), "PS=Yes", "PS=No")
```

###### adj PC2

```{r, results='asis'}
adjPC2 = residuals(lm(PC2 ~ sex + age, data=patho_df))
x = adjPC2[patho_df$PS == "Yes"]
y = adjPC2[patho_df$PS == "No"]
report.t(apa::t_test(x, y), "PS=Yes", "PS=No")
```

```{r}
## Plot PS
patho_df %>% select(PC1, PC2, PS) %>% 
  ggplot(aes(x=PC1, y=PC2, color = PS)) + 
  geom_point() + 
  geom_point(data = patho_df %>% 
               select(PC1, PC2, PS) %>% 
               group_by(PS) %>%  
               summarize_all(mean), 
             size = 4, shape = 17) + 
  scale_color_manual(values = c("magenta", "gold")) + 
  ggtitle("22q-like PC scores for 22q sample coded by PS")
```

##### DSM Diagnosis

```{r, include=FALSE}
## Plot DMS Dx
patho_df %>% 
  ggplot(aes(x=PC1, y=PC2, color = `DSM Dx`)) + 
  geom_point() + 
  geom_point(data = patho_df %>% select(PC1, PC2, `DSM Dx`) %>% 
               group_by(`DSM Dx`) %>%  
               summarize_all(mean), 
             size = 4, shape = 17) + 
scale_color_manual(values = c("magenta", "gold")) + 
  ggtitle("22q-like PC scores for 22q sample coded by DSM Dx")
```

There was no effect of having a confirmed DSM diagnosis on PC2 scores

###### PC2

```{r, results='asis'}
x = patho_df$PC2[patho_df$`DSM Dx` == "Yes"]
y = patho_df$PC2[patho_df$`DSM Dx` == "No"]
report.t(apa::t_test(x, y), "DSM=Yes", "DSM=No")
```

###### adj PC2

```{r, results='asis'}
x = adjPC2[patho_df$`DSM Dx` == "Yes"]
y = adjPC2[patho_df$`DSM Dx` == "No"]
report.t(apa::t_test(x, y), "DSM=Yes", "DSM=No")
```

#### Emotrics

PC2 scores were correlated with Brow Height (Left and Right), Marginal Reflex Distance 2 (Left and Right), and Philtrum measurements. After adjustment, however, all correlations with Emotrics are significant.

```{r}
emotrics = c('Brow_Height_Right', 'Brow_Height_Left', 'Marginal_Reflex_Distance_1_Right', 'Marginal_Reflex_Distance_1_Left', 'Marginal_Reflex_Distance_2_Right', 'Marginal_Reflex_Distance_2_Left', 'Philtrum')

emotrics_df = dat %>% select(bbl_id, case_id, all_of(emotrics)) %>% mutate(Brow_Height = (Brow_Height_Right + Brow_Height_Left) / 2, Marginal_Reflex_Distance_1 = (Marginal_Reflex_Distance_1_Left + Marginal_Reflex_Distance_1_Right) /2, Marginal_Reflex_Distance_2 = (Marginal_Reflex_Distance_2_Right + Marginal_Reflex_Distance_2_Left)/2) %>% select(-ends_with('Left'), -ends_with('Right'))

emotrics = colnames(emotrics_df)[-c(1:2)]
  
# Add emotrics to scores data.frame
emo_df = merge(cov_df, emotrics_df, by = 'case_id') %>% drop_na()
res.cor = psych::corr.test(emo_df[, "PC2"],
                           emo_df[, emotrics], 
                           adjust='fdr')

cor.mat = rbind(res.cor$r[1,], 
                res.cor$p.adj[1,])

rownames(cor.mat) = c('PC2', 'p')

emo_df$PC2 = residuals(lm(PC2 ~ sex + age, data=emo_df)) 
res.cor = psych::corr.test(emo_df[, "PC2"],
                           emo_df[, emotrics], 
                           adjust='fdr')
cor.mat = rbind(cor.mat,
                res.cor$r[1,], 
                res.cor$p.adj[1,])

rownames(cor.mat)[3:4] = c('adj PC2', 'p')
cor.mat %>% t() %>% kable(format = 'html') %>%  kable_styling("striped", full_width = TRUE)
sig.emotrics = names(which(cor.mat[2,] < .05))
```

#### IQ

There were no significant relationships with IQ.

```{r}
iq = c("VIQ", "PIQ", "FSIQ")
iq_df = dat %>% select(case_id, all_of(iq)) 

# Add emotrics to scores data.frame
iq_df = merge(cov_df, iq_df, by = 'case_id') %>% drop_na()
res.cor = psych::corr.test(iq_df[, "PC2"], 
                       iq_df[, iq], 
                       adjust='fdr')

cor.mat = rbind(res.cor$r[1,], 
                res.cor$p.adj[1,])

rownames(cor.mat) = c('PC2', 'p')
# Adjust scores
iq_df$PC2 = residuals(lm(PC2 ~ sex + age, data=iq_df)) 
res.cor = psych::corr.test(iq_df[, "PC2"],
                           iq_df[, iq], 
                           adjust='fdr')
cor.mat = rbind(cor.mat,
                res.cor$r[1,], 
                res.cor$p.adj[1,])

rownames(cor.mat)[3:4] = c('adj PC2', 'p')
cor.mat %>% t() %>% kable(format = 'html') %>%  kable_styling("striped", full_width = TRUE)
ssig.iq = names(which(cor.mat[4,] < .05))
```

#### Height

There were no significant relationships with height and weight.

```{r}
height_df = dat %>% select(case_id, all_of(c("height", "weight")))

# Add emotrics to scores data.frame
height_df$height = sapply(height_df$height,
                          FUN = function(x) {
                            x = unlist(strsplit(x, "'"))
                            ft = as.numeric(x[1]) * 12
                            inch = as.numeric(x[2])
                            ft + inch
                          })
height_df = merge(cov_df, height_df, by = 'case_id') %>% drop_na()
res.cor = psych::corr.test(height_df[, "PC2"], 
                           height_df[, c("height", "weight")],
                           adjust='fdr')

cor.mat = rbind(res.cor$r[1,], 
                res.cor$p.adj[1,])

rownames(cor.mat) = c('PC2', 'p')

height_df$PC2 = 

# Adjust scores
height_df$PC2 = residuals(lm(PC2 ~ sex + age, data=height_df)) 
res.cor = psych::corr.test(height_df[, "PC2"],
                           height_df[, c("height", "weight")], 
                           adjust='fdr')
cor.mat = rbind(cor.mat,
                res.cor$r[1,], 
                res.cor$p.adj[1,])

rownames(cor.mat)[3:4] = c('adj PC2', 'p')


cor.mat %>% t() %>% kable(format = 'html') %>%  kable_styling("striped", full_width = TRUE)
sig.height = names(which(cor.mat[4,] < .05))
```

#### GAF

There were no significant relationships with GAF.

```{r}
gaf = c("GAF_C", "GAF_H")
gaf_df = dat %>% select(case_id, all_of(gaf)) 

# Add emotrics to scores data.frame
gaf_df = merge(cov_df, gaf_df, by = 'case_id') %>% drop_na()
res.cor = psych::corr.test(gaf_df[, "PC2"], 
                       gaf_df[, gaf], 
                       adjust='fdr')

cor.mat = rbind(res.cor$r[1,], 
                res.cor$p.adj[1,])

rownames(cor.mat) = c('PC2', 'p')
# Adjust PC2
gaf_df$PC2 = residuals(lm(PC2 ~ sex + age, data=gaf_df))
res.cor = psych::corr.test(gaf_df[, "PC2"], 
                       gaf_df[, gaf], 
                       adjust='fdr')
cor.mat = rbind(cor.mat,
                res.cor$r[1,], 
                res.cor$p.adj[1,])

rownames(cor.mat)[3:4] = c('adj PC2', 'p')

cor.mat %>% t() %>% kable(format = 'html') %>% kable_styling("striped", full_width = TRUE)
sig.gaf = names(which(cor.mat[4,] < .05))
```

#### MMSE

There was a significant negative relationship between PC2 and MMSE scores.

```{r}
mmse_df = dat %>% select(case_id, all_of("MMSE")) 

# Add emotrics to scores data.frame
mmse_df = merge(cov_df, mmse_df, by = 'case_id') %>% drop_na()
res.cor = psych::corr.test(mmse_df[, "PC2"], 
                           mmse_df[, "MMSE"],
                           adjust='fdr')

cor.mat = rbind(res.cor$r[1], 
                res.cor$p.adj[1])

rownames(cor.mat) = c('PC2', 'p')

mmse_df$PC2 = residuals(lm(PC2 ~ sex + age, data=mmse_df))
res.cor = psych::corr.test(mmse_df[, "PC2"], 
                           mmse_df[, "MMSE"],
                           adjust='fdr')

cor.mat = rbind(cor.mat,
                res.cor$r[1], 
                res.cor$p.adj[1])
rownames(cor.mat)[3:4] = c('adj PC2', 'p')

cor.mat %>% t() %>% kable(format = 'html') %>%  kable_styling("striped", full_width = TRUE)
sig.mmse = names(which(cor.mat[4,] < .05))
```

## 2. Testing for differences between TD and PS

### 22q-like factor scores

For the Penn sample (i.e. PS and TD groups), 2 factor scores are computed using the loadings shown above as weights in a linear combination of the relevant gestalt scores.

**Note:** any reference to PCs/factor scores from this point refers to those computed for the Penn sample. 

```{r}
weigh_scores = function(df, coefs){
  #' Returns a linear combination of named scores included in 'coefs' 
  #' If named scores in coefs are not defined in df, columns are initialized with 0
  
  # which gestalt scores are missing from penn_df? i.e. PS and TD groups
  nohits = names(coefs)[which(!(names(coefs) %in% colnames(df)))] 
  df[, nohits] = 0 # set missing to zero to define columns
  
  # linear combination of gestalt scores
  scores.mat = as.matrix(df[, names(coefs)])
  weighted_score = as.vector(scores.mat %*% coefs)
  return(weighted_score)
}

weigh_scores2 = function(df, resPCA){
  # which gestalt scores are missing from penn_df? i.e. PS and TD groups
  nohits = rownames(resPCA$var$coord)[which(!(rownames(resPCA$var$coord) %in% colnames(df)))] 
  df[, nohits] = 0 # set missing to zero to define columns
   
  scores.mat = as.matrix(df[, rownames(resPCA$var$coord)]) 
  weighted_score = scores.mat %*% resPCA$svd$V
  
  return(weighted_score)
}
```



```{r, include=FALSE, eval=FALSE}
#### PC 1

#No significant differences between groups.



res.t.test = t.test(scores_df$PC1[TD], scores_df$PC1[PS])
names(res.t.test$estimate) = c("mean of TD", "mean of PS")
res.t.test
```

#### TD vs PS

##### PC2

PC2 shows a difference between TD and PS groups.

```{r, results='asis'}
scores_df = data.frame(group = penn_df$group)
scores_df$PC1 = weigh_scores(penn_df[, which(colnames(penn_df) == '22q11.2 deletion syndrome'):ncol(penn_df)],
                             get_loadings(resPCA, 1))
# TD = df$group == "TD"
TD = scores_df$group == 'TD'
PS = scores_df$group == "PS"


scores_df$PC2 = weigh_scores(penn_df[, which(colnames(penn_df) == '22q11.2 deletion syndrome'):ncol(penn_df)], get_loadings(resPCA, 2))

# bartlett.test(scores_df$PC2, scores_df$group) # Can assume equalvariance
res.t = apa::t_test(scores_df$PC2[TD], scores_df$PC2[PS], var.equal=TRUE)
report.t(res.t, "TD", "PS")
```

```{r}
scores_df %>% 
  ggplot(aes(x=PC1, y=PC2, color = group)) + 
  geom_point() + 
  geom_point(data = scores_df %>% group_by(group) %>% 
               summarize_all(mean), 
             size = 4, shape = 17) + 
  scale_color_manual(values = c("salmon", "dodgerblue")) + 
  ggtitle("22q-like PC Scores of Penn sample")
```

##### adj PC2

```{r, results='asis'}
cov_df = data.frame(scores_df, 
                    penn_df[, c("case_id", "sex", "age", "race")])

adjPC2 = residuals(lm(PC2 ~ sex + age, data=cov_df))
res.t = apa::t_test(adjPC2[cov_df$group == "TD"], adjPC2[cov_df$group == "PS"], var.equal=TRUE)
report.t(res.t, "TD", "PS")
```

#### TD vs SZ

Differences remain when comparing TD only to SZ. 

##### PC2

```{r, results='asis'}
sz_df = merge(sz_cr, data.frame(penn_df[, c("bbl_id", "sex", "age")], select(scores_df, !group)), by="bbl_id") %>% filter(group != 'CR') %>% drop_na()

res.t = apa::t_test(sz_df[sz_df$group == "TD", "PC2"], sz_df[sz_df$group == "SZ", "PC2"], var.equal=TRUE)
report.t(res.t, "TD", "SZ")
```

There are no variance difference between these two groups.

```{r}
bartlett.test(sz_df$PC2, sz_df$group) # Can assume equal variance
```

##### adjPC2

```{r, results='asis'}
adjPC2 = residuals(lm(PC2 ~ sex + age, data=sz_df))
res.t = apa::t_test(adjPC2 [sz_df$group == "TD"], adjPC2 [sz_df$group == "SZ"], var.equal=TRUE)
report.t(res.t, "TD", "SZ")
```

There are no variance difference between these two groups.

```{r}
bartlett.test(adjPC2, sz_df$group) # Can assume equal variance
```

#### TD vs CR

Comparison of TD with CR shows there are no differences in means

##### PC2

```{r, results='asis'}
cr_df = merge(sz_cr, data.frame(penn_df[, c("bbl_id", "sex", "age")], select(scores_df, !group)), by="bbl_id") %>% filter(group != 'SZ') %>% drop_na()


res.t = apa::t_test(cr_df[cr_df$group == "TD", "PC2"], cr_df[cr_df$group == "CR", "PC2"], var.equal=TRUE)
report.t(res.t, "TD", "CR")
```

But there are variance differences. 

```{r}
bartlett.test(cr_df$PC2, cr_df$group) # Can assume equal variance
```




##### adjPC2

```{r, results='asis'}
adjPC2 = residuals(lm(PC2 ~ sex + age, data=cr_df))

res.t = apa::t_test(adjPC2[cr_df$group == "TD"], adjPC2[cr_df$group == "CR"], var.equal=FALSE)
report.t(res.t, "TD", "SZ")
```

#### SZ vs CR

##### PC2

No differences in means.

```{r}
sz_cr_df = merge(sz_cr, data.frame(penn_df[, c("bbl_id", "sex", "age")], select(scores_df, !group), by="bbl_id")) %>% filter(group != 'TD') %>% drop_na()
```


```{r, results='asis'}
res.t = apa::t_test(sz_cr_df[sz_cr_df$group == "SZ", "PC2"], sz_cr_df[sz_cr_df$group == "CR", "PC2"], var.equal=TRUE)
report.t(res.t, "SZ", "CR")
```

But We see a difference in variances on PC2.
 
```{r}
bartlett.test(sz_cr_df$PC2, sz_cr_df$group)
```

 
```{r}
sz_cr_df %>% 
  ggplot(aes(x=PC1, y=PC2, color = group)) + 
  geom_point() + 
  geom_point(data = sz_cr_df %>%  
               group_by(group) %>% 
               summarize_all(mean), 
             size = 4, shape = 17) + 
  scale_color_manual(values = c("orange", "purple"))
```

##### adjPC2

Results remain after adjustment. 

```{r, results='asis'}
adjPC2 = residuals(lm(PC2 ~ sex + age, data=sz_cr_df))
res.t = apa::t_test(adjPC2[sz_cr_df$group == "SZ"], adjPC2[sz_cr_df$group == "CR"], var.equal=TRUE)
report.t(res.t, "SZ", "CR")
```
 
```{r}
bartlett.test(adjPC2, sz_cr_df$group)
```


### Emotrics

```{r}
# Order emotrics 
emo_df = merge(data.frame(case_id = penn_df$case_id,
                          scores_df), 
               emotrics_df, 
               by='case_id')
# emo_df %>%  select(all_of(emotrics)) %>% lapply(bartlett.test, g=emo_df$group) # All variances are equal after adj



res.t = emo_df %>% 
  select(all_of(emotrics)) %>% 
  lapply(function(x) apa::t_test(x[emo_df$group == 'PS'], x[emo_df$group == 'TD'], 
                                 var.equal=TRUE))

report.t.list = function(res.t){
  #' Report t-test in apa format
  #'' TODO: to fully automate: print groups from data.name, bonus: add formatting/substutions of var.name as parameter
    for(dep.var in names(res.t)){
      
        cat(sprintf("- %s: PS (*M* = %s, *SD* = %s) and TD (*M* = %s, *SD* = %s), %s", 
                gsub("Right", "(R)", gsub("Left", "(L)", gsub("_", " ", dep.var))),
                round(mean(res.t[[dep.var]]$data$x), 2), round(sd(res.t[[dep.var]]$data$x), 2),
                round(mean(res.t[[dep.var]]$data$y), 2), round(sd(res.t[[dep.var]]$data$y), 2),
                apa::apa(res.t[[dep.var]], print = FALSE)), 
            sep="\n")
    }
}
```

For univariate t-tests, the only difference between groups was in Philtrum measurements and Marginal Reflex Distance 1:  

```{r, results='asis'}
report.t.list(res.t)
```

### GAF_C

GAF_C **are** significantly different between PS and TD.

```{r, results='asis'}
ps_td_gafc = dat %>% 
  select(group, GAF_C) %>% filter(group !='22q')

apa::t_test(ps_td_gafc$GAF_C[ps_td_gafc$group == 'TD'],
            ps_td_gafc$GAF_C[ps_td_gafc$group == 'PS'], var.equal=TRUE) %>% report.t('TD', 'PS')
```


### GAF_H

GAF_H **are** significantly different between PS and TD.

```{r, results='asis'}
ps_td_gafh = dat %>% 
  select(group, GAF_H)

apa::t_test(ps_td_gafh$GAF_H[ps_td_gafh$group == 'TD'],
       ps_td_gafh$GAF_H[ps_td_gafh$group == 'PS'], var.equal=TRUE) %>% report.t('TD', 'PS')
```

### Height

No significant difference.

```{r, results='asis'}
ps_td_height = dat %>% 
  select(group, height)

ps_td_height$height = sapply(ps_td_height$height,
                          FUN = function(x) {
                            x = unlist(strsplit(x, "'"))
                            ft = as.numeric(x[1]) * 12
                            inch = as.numeric(x[2])
                            ft + inch
                          })
apa::t_test(ps_td_height$height[ps_td_height$group == 'TD'],
       ps_td_height$height[ps_td_height$group == 'PS'], var.equal=TRUE) %>% report.t(x='TD', y='PS')
```

### Weight

No significant difference.

```{r, results='asis'}
ps_td_weight = dat %>% 
  select(group, weight)

apa::t_test(ps_td_weight$weight[ps_td_weight$group == 'TD'],
            ps_td_weight$weight[ps_td_weight$group == 'PS'], var.equal=TRUE) %>% report.t(x='TD', y='PS')
```


### MMSE

No significant difference.

```{r, results='asis'}
ps_td_mmse = dat %>% 
  select(group, MMSE)

apa::t_test(ps_td_mmse$MMSE[ps_td_mmse$group == 'TD'],
       ps_td_mmse$MMSE[ps_td_mmse$group == 'PS'], var.equal=TRUE) %>% report.t('TD', 'PS')
```

## 3. Predicting PS status based on 22q-like PC2

```{r}
plotROCs = function(roc1, roc2, roc3, main, legends){
  plot(roc1, col="black", main=main)
  plot(roc2, col="purple", add=TRUE)
  plot(roc3, col="orange", add=TRUE)
  
  AUC = c(round(as.vector(roc1$auc), 2),
          round(as.vector(roc2$auc), 2),
          round(as.vector(roc3$auc), 2))
  
  
  legend("bottomright", 
    legend = paste0(legends, " (", AUC, ")"),
    col = c("black", "purple", "orange"), 
    pch = c(17,19, 19), 
    pt.cex = 1, 
    cex = 1, 
    text.col = "black", 
    inset = c(0.1, 0.1))
}
```

```{r, message=FALSE, results='asis'}
resROC_main_ps = roc(group ~ PC2, data=scores_df)

resROC_main_sz = roc(group ~ PC2, data=sz_df)

resROC_main_cr = roc(group ~ PC2, data=cr_df)

legends = c('PS', 'SZ', 'CR')
plotROCs(resROC_main_ps, resROC_main_sz, resROC_main_cr, main = 'ROC Curves of PC2 predicting Status (AUC)', legends =  legends)
```

#### Psychosis Status

```{r, results='asis'}
ci(resROC_main_ps)
```

#### Schizophrenia Status 

```{r, results='asis', }
ci(resROC_main_sz)
```

#### Clinical Risk Status

```{r, results='asis'}
ci(resROC_main_cr)
```

## 3b. Predicting PS status based on Emotrics

Prediction by the 4 (aggregated) Emotrics variables is significant for PS and SZ status. Relative to prediction by PC2, prediction by emotrics is a slightly weaker -- though confidence intervals overlap.

```{r, message=FALSE, results='asis'}

loo.log = function(df){ # Leave one out logistic regression with emotrics as predictor
  df$group = as.numeric(df$group != 'TD')
  frmla = as.formula(sprintf("group ~ %s", paste(emotrics, collapse=" + ")))
  reslog = c()
  for (i in 1:nrow(df)){ ## Leave one out CV
    tmp_df = df[-i,]
    reslog[i] = glm(frmla, family='binomial', data=tmp_df)$fitted.values
  }
  return(reslog)
}

emo_df$reslog = loo.log(emo_df)
resROC_main_ps = roc(group ~ reslog, data=emo_df)

sz_df = merge(sz_df, emotrics_df, by='bbl_id')
sz_df$reslog = loo.log(sz_df)
resROC_main_sz = roc(group ~ reslog, data=sz_df)

cr_df = merge(cr_df, emotrics_df, by='bbl_id')
cr_df$reslog = loo.log(cr_df)
resROC_main_cr = roc(group ~ reslog, data=cr_df)

legends = c('PS', 'SZ', 'CR')
plotROCs(resROC_main_ps, resROC_main_sz, resROC_main_cr, main = 'ROC Curves of Emotrics predicting Status (AUC)', legends =  legends)
```

#### Psychosis Status

```{r, results='asis'}
ci(resROC_main_ps)
```

#### Schizophrenia Status 

```{r, results='asis', }
ci(resROC_main_sz)
```

#### Clinical Risk Status

```{r, results='asis'}
ci(resROC_main_cr)
```

## 4. Predicting PS status based on adjusted PC2 

### Sex and age adjustment

Prediction strength for PS status drops slightly after adjustment. However, prediction of CR status becomes essentially random (see CI).

```{r, message=FALSE}
cov_df = data.frame(scores_df, 
                    penn_df[, c("case_id", "sex", "age", "race")]) %>%
  drop_na()
cov_df$PC2 = residuals(lm(PC2 ~ sex + age, data=cov_df))
resROC_ps = roc(group ~ PC2, data=cov_df)

sz_cr_df = data.frame(bbl_id = penn_df$bbl_id,
                   group = penn_df$group,
                   age = penn_df$age) %>%
  drop_na()

sz_cr_df$PC2 = cov_df$PC2

sz_cr_df = merge(sz_cr, sz_cr_df %>% select(!group),
              by = 'bbl_id') 
sz_df = sz_cr_df %>%
  filter(group != 'CR')

resROC_sz = roc(group ~ PC2, data=sz_df)

cr_df = sz_cr_df %>%
  filter(group != 'SZ')

resROC_cr = roc(group ~ PC2, data=cr_df)
plotROCs(resROC_ps, resROC_sz, resROC_cr, main = 'ROC Curves of PC2 predicting Status (AUC)',legends =  legends)
```

#### Psychosis Status

```{r, results='asis'}
ci(resROC_ps)
```

#### Schizophrenia Status 

```{r, results='asis', }
ci(resROC_sz)
```

#### Clinical Risk Status

```{r, results='asis'}
ci(resROC_cr)
```

### Emotrics-adjusted scores

After adjusting PC2 for emotrics, prediction of SZ-status remains strong while CR becomes random.

```{r, message=FALSE}
emo_df$PC2 = residuals(lm(formula(sprintf("%s ~ %s", 
                                          "PC2", 
                                          paste(sig.emotrics, collapse=" + "))), 
                          data=emo_df))
resROC_ps = roc(group ~ PC2, data=emo_df)

sz_df = merge(sz_cr, 
              emo_df[,c("bbl_id", "PC2", emotrics)],) %>% 
  filter(group != 'CR')
resROC_sz = roc(group ~ PC2, data=sz_df)

cr_df = merge(sz_cr, 
              emo_df[,c("bbl_id", "PC2", emotrics)]) %>% 
  filter(group != 'SZ')
resROC_cr = roc(group ~ PC2, data=cr_df)
plotROCs(resROC_ps, resROC_sz, resROC_cr, main = 'ROC Curves of PC2 predicting Status (AUC)',legends =  legends)
```

#### Psychosis Status

```{r, results='asis'}
ci(resROC_ps)
```

#### Schizophrenia Status 

```{r, results='asis', }
ci(resROC_sz)
```

#### Clinical Risk Status

```{r, results='asis'}
ci(resROC_cr)
```

### High missingness drops AUC

#### GAF_C

Excluding observations with missing data on GAF_C, drops AUC to random levels (i.e. CI includes .50) pointing to insufficient data. No adjustment was applied to PC2 scores.

```{r}
t(n_tbl["GAF_C",]) %>% kable(format="html") %>% kable_styling("striped", full_width = TRUE)
```

```{r, message=FALSE}
cov_gafc_df = data.frame(scores_df, 
                    penn_df[, c("case_id", "bbl_id", "GAF_C")]) %>% drop_na()
#cov_gafc_df$PC2 = residuals(lm(PC2 ~ GAF_C, data=cov_gafc_df))
resROC_ps = roc(group ~ PC2, data=cov_gafc_df)

sz_gafc_df = merge(cov_gafc_df %>%
                       select(!group),
                     sz_cr,
                     by="bbl_id") %>%
  filter(group != 'CR')

resROC_sz = roc(group ~ PC2, data=sz_gafc_df)

cr_gafc_df = merge(cov_gafc_df %>%
                       select(!group),
                     sz_cr,
                     by="bbl_id") %>%
  filter(group != 'SZ')


resROC_cr = roc(group ~ PC2, data=cr_gafc_df)
plotROCs(resROC_ps, resROC_sz, resROC_cr, main = 'ROC Curves of PC2 predicting Status (AUC)',legends =  legends)
```

##### Psychosis Status

```{r, results='asis'}
ci(resROC_ps)
```

##### Schizophrenia Status 

```{r, results='asis', }
ci(resROC_sz)
```

##### Clinical Risk Status

```{r, results='asis'}
ci(resROC_cr)
```

#### GAF_H

Excluding observations with missing data on GAF_H, drops AUC. No adjustment was applied to PC2 scores.
```{r}
t(n_tbl["GAF_C",]) %>% kable(format="html") %>% kable_styling("striped", full_width = TRUE)
```


```{r, message=FALSE}
cov_gafh_df = data.frame(scores_df, 
                    penn_df[, c("case_id", "bbl_id", "GAF_H")])  %>% drop_na()
#cov_gafh_df$PC2 = residuals(lm(PC2 ~ GAF_H, data=cov_gafh_df))
resROC_ps = roc(group ~ PC2, data=cov_gafh_df)

sz_gafh_df = merge(cov_gafh_df %>%
                       select(!group),
                     sz_cr,
                     by="bbl_id") %>%
  filter(group != 'CR')

resROC_sz = roc(group ~ PC2, data=sz_gafh_df)

cr_gafh_df = merge(cov_gafh_df %>%
                       select(!group),
                     sz_cr,
                     by="bbl_id") %>%
  filter(group != 'SZ')


resROC_cr = roc(group ~ PC2, data=cr_gafh_df)
plotROCs(resROC_ps, resROC_sz, resROC_cr, main = 'ROC Curves of PC2 predicting Status (AUC)',legends =  legends)
```

##### Psychosis Status

```{r, results='asis'}
ci(resROC_ps)
```

##### Schizophrenia Status 

```{r, results='asis', }
ci(resROC_sz)
```

##### Clinical Risk Status

```{r, results='asis'}
ci(resROC_cr)
```

#### MMSE

Excludding observations with missing data on MMSE. No adjustment was applied to PC2 scores.

```{r}
t(n_tbl["MMSE",]) %>% kable(format="html") %>% kable_styling("striped", full_width = TRUE)
```

```{r, message=FALSE}
cov_mmse_df = data.frame(scores_df, 
                    penn_df[, c("case_id", "bbl_id", "MMSE")]) %>%
  drop_na()

resROC_ps = roc(group ~ PC2, data=cov_mmse_df)
#cov_mmse_df$PC2 = residuals(lm(PC2 ~ MMSE, data=cov_mmse_df))

sz_mmse_df = merge(cov_mmse_df %>%
                       select(!group),
                     sz_cr,
                     by="bbl_id") %>%
  filter(group != 'CR')

resROC_sz = roc(group ~ PC2, data=sz_mmse_df)

cr_mmse_df = merge(cov_mmse_df %>%
                       select(!group),
                     sz_cr,
                     by="bbl_id") %>%
  filter(group != 'SZ')


resROC_cr = roc(group ~ PC2, data=cr_mmse_df)
plotROCs(resROC_ps, resROC_sz, resROC_cr, main = 'ROC Curves of PC2 predicting Status (AUC)',legends =  legends)
```

##### Psychosis Status

```{r, results='asis'}
ci(resROC_ps)
```

##### Schizophrenia Status 

```{r, results='asis', }
ci(resROC_sz)
```

##### Clinical Risk Status

```{r, results='asis'}
ci(resROC_cr)
```

## Conclusion

- PC2 alone is predictive of PS status with an AUC of `r round(pROC::auc(resROC_main_ps), 3)`!
- There are variance differences on PC2 between PS groups (SZ vs CR).
- By examining scores from 22q sample, PC2 can be interpreted as showing an association with Psychosis Symptoms, Face measurements, and MMSE scores, with age and sex effects.
- After adjustments, Schizophrenia (SZ) status remains well-predicted while Clinical Risk (CR) status does not.
