---
title: "Face2Gene Exploratory Analysis"
author: "Virgilio Gonzenbach"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document:
    toc: true
    toc_float: true
    theme: spacelab

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      cache = TRUE, 
                      cache.path = 'cache/ROC+PCA/', 
                      fig.path = "figures/ROC+PCA/")
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
```



```{r load, include=FALSE}
library(dplyr)
library(tidyr)
library(kableExtra)
library(ROCR)
library(FactoMineR)
library(factoextra)
library(psych)
library(pROC)

# Load data
penn_df = read.csv("data/Penn_merged_wideV.csv")[,-1] %>% filter(group != "?") # 3 observations in '?' group
chop_df = read.csv('data/Chop_merged_wideV_tp1.csv')

# Exclude columns with only zeros
get_zero_cols = function(df){
  num_zeros = df %>% sapply(function(x) x == 0) %>% colSums()
  prop_zeros = num_zeros / nrow(chop_df)
  zero_cols = names(which(prop_zeros == 1))
  return(zero_cols)
}

chop_df = chop_df %>% select(-all_of(get_zero_cols(chop_df)))

# Merge data
df = merge(penn_df, chop_df, all = TRUE, sort = FALSE)
df[ , c("age_at_photo", "visit_id")] = NULL

# Fill missing values with 0
df[is.na.data.frame(df)]  = 0

# Set seed for reproducibility
set.seed(42)
```

## ROC analysis

The ROC analysis presented here was performed on the data including all groups (22q, CR, ZR, NC). Groups status was coded as "22q" or "not22q". 

Sample sizes:

```{r}
# Print sample sizes
t(table(df$group)) %>% kable(format="html") %>% kable_styling("striped", full_width = TRUE)
```


```{r}
# Plot 22q gestalt score values per group
df %>% ggplot(aes(x=group, y=X22q11.2.deletion.syndrome, color=group)) + geom_boxplot() + ggtitle("Group means of 22q11.2 deletion syndrome gestalt score")
```

```{r run_roc, include=FALSE}
# Prep data for ROC
roc_df = df %>% select(group, X22q11.2.deletion.syndrome)
roc_df$group = ifelse(df$group == '22q', 1, 0) 

# Run ROC
resROC = roc(group ~ X22q11.2.deletion.syndrome, data=roc_df)
```

### ROC curve

ROC curve with optimal cut-off point (Specificity, Sensitivity). Optimal cut-off is defined as the point that maximizes Youden's index defined as $J = sensitivity + specificity − 1$

```{r print_roc}
# Plot ROC results
plot(resROC, main="ROC curve of 22q vs not 22q (whole sample)", print.thres="best", print.thres.best.method="youden")
auc(resROC)
ci.auc(resROC)
```

## PCA

PCA was performed without scaling gestalt scores (covariance PCA).

### Whole sample

#### Scree

```{r}
resPCA = PCA(df[,-c(1:6)], scale.unit = FALSE, graph = FALSE)
fviz_screeplot(resPCA, barcolor = "blueviolet", barfill="blueviolet")
```

#### Factor scores

```{r}
fviz_pca_ind(resPCA, geom = c("point"), col.ind = df$group)
```

```{r}
fviz_pca_ind(resPCA, axes = c(3,4), geom = c("point"), col.ind = df$group)
```

#### Parallel Analysis

Needs work.

```{r}
plot_loading = function(resPCA, dim=1){
  
  # Extract metrics from PCA object
  r = resPCA$var$coord[,dim] / sqrt(resPCA$eig[dim,1])

  filter = r^2 > mean(r^2, na.rm = TRUE) # only include loadings with more than average contributions
  
  # Prep data
  df = data.frame(var=names(r), cor = r, row.names = NULL)[filter, ] %>% drop_na()
  df = df[order(df$cor, decreasing = TRUE),]
  df$var = factor(df$var, levels = rev(df$var)) #order before plotting
  
  # Plot
  title = sprintf("PC%s contributions", dim)
  p = df %>% ggplot(aes(x=var, y=cor)) + geom_col() + coord_flip() + 
    xlab("Gestalt Score") + ylab("Loading") + ggtitle(title) + 
    theme_bw() + theme(axis.text.y = element_text(size=10))
  return(p)
}
```

#### PC1

##### Loadings

```{r, fig.width=12, fig.height=8}
plot_loading(resPCA, dim = 1)
```

##### ROC Analysis 

```{r boxplot_PC1}
# Run ROC
roc_df = data.frame(score = resPCA$ind$coord[,1], group = df$group)
roc_df %>% ggplot(aes(x=group, y=score, color=group)) + geom_boxplot() + ggtitle("Group means of PC1 score") + ylab("PC1 score")
```


```{r run_roc2, include=FALSE}
roc_df$group = ifelse(df$group == '22q', 1, 0)
resROC_PCA = roc(group ~ score, data = roc_df)
```


```{r plot_roc2}
plot(resROC_PCA, main = "ROC curve of 22q vs not with PC1", print.thres="best", print.thres.best.method="youden")
auc(resROC_PCA)
ci.auc(resROC_PCA)
```

##### Compare ROCs: 22q Gestalt score vs PC1 as input

```{r delong}
plot(resROC, col="blue")
plot(resROC_PCA, col="red", add=TRUE)
legend("bottomright", 
  legend = c("22q Gestalt Score", "PC1"), 
  col = c("blue", "red"), 
  pch = c(17,19), 
  bty = "n", 
  pt.cex = 2, 
  cex = 1.2, 
  text.col = "black", 
  horiz = F , 
  inset = c(0.1, 0.1))
roc.test(resROC, resROC_PCA, method = "delong")
```

A weighted combination of syndrome scores improves the AUC relative to predicting 22q status based on only 22q syndrome scores. 

#### PC2

##### Loadings

```{r, fig.width=12, fig.height=8}
plot_loading(resPCA, dim = 2)
```

##### ROC Analysis

```{r boxplot_pc2}
# Run ROC
roc_df = data.frame(PC2 = resPCA$ind$coord[df$group!="22q", 2])
roc_df$group = df$group[df$group != "22q"]
roc_df %>% ggplot(aes(x=group, y=PC2, color=group)) + geom_boxplot() + ggtitle("Group means of PC2 score") + ylab("PC2 score")
```


```{r run_roc3, include=FALSE}
roc_df$group = ifelse(roc_df$group != 'NC', 1, 0)
resROC_PC2 = roc(group ~ PC2, data = roc_df)
```


```{r plot_roc3}
plot(resROC_PC2, main = "ROC curve of PS vs NC with PC2", print.thres="best", print.thres.best.method="youden")
auc(resROC_PC2)
ci.auc(resROC_PC2)
```

###### T-test: Normal Controls vs Schizophrenia

```{r}
NC_score = resPCA$ind$coord[df$group == "NC", 2]
SZ_score = resPCA$ind$coord[df$group == "SZ", 2]
t.test(NC_score, SZ_score)
effectsize::cohens_d(NC_score, SZ_score)
```

###### T-test: Normal Controls vs Clinical Risk

```{r}
CR_score = resPCA$ind$coord[df$group == "CR", 2]
t.test(NC_score, CR_score)
effectsize::cohens_d(NC_score, CR_score)
```

###### T-test: Schizophrenia vs Clinical Risk

```{r}
t.test(SZ_score, CR_score)
effectsize::cohens_d(SZ_score, CR_score)
```

#### PC3

##### Loadings

```{r, fig.width=12, fig.height=8}
plot_loading(resPCA, dim = 3)
```

```{r}
data.frame(score = resPCA$ind$coord[,3], group = df$group) %>% ggplot(aes(x=group, y=score, color=group)) + geom_boxplot() + ggtitle("Group means of PC3 score") + ylab("PC3 score")
```

#### PC4

##### Loadings

```{r, fig.width=12, fig.height=8}
plot_loading(resPCA, dim = 4)
```

##### Group means

```{r}
data.frame(score = resPCA$ind$coord[,4], group = df$group) %>% ggplot(aes(x=group, y=score, color=group)) + geom_boxplot() + ggtitle("Group means of PC4 score") + ylab("PC4 score")
```

### 22q group

#### Scree

```{r}
resPCA = PCA(chop_df[,-c(1:7)], scale.unit = FALSE, graph = FALSE)
fviz_screeplot(resPCA)
```

#### Loadings

##### PC1

```{r, fig.width=12, fig.height=8}
plot_loading(resPCA, dim=1)
```

##### PC2

```{r, fig.width=12, fig.height=8}
plot_loading(resPCA, dim=2)
```

##### PC3

```{r, fig.width=12, fig.height=8}
plot_loading(resPCA, dim=3)
```

##### PC4

```{r, fig.width=12, fig.height=8}
plot_loading(resPCA, dim=4)
```